---
title: Functions and testing
teaching: 45
exercises: 45
questions:
- "Functions:"
- "How can I write a new function in R?"
- "Testing:"
- "How can I test my functions"
objectives:
- "Functions:"
- "Define a function that takes arguments."
- "Return a value from a function."
- "Test a function."
- "Set default values for function arguments."
- "Explain why we should divide programs into small, single-purpose functions."
- "Testing:"
- "Understanding the necessity for testing"
- "Making __testing__ more fun"
keypoints:
- "Functions:"
- "Use `function` to define a new function in R."
- "Use parameters to pass values into functions."
- "Load functions into programs using `source`."
- "Testing:"
- "write `tests` instead of `print` statements"

source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02_session-")
# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("data/gapminder-FiveYearData.csv", header=TRUE)
```

In this lesson, we'll learn how to write a function so that we can repeat
several operations with a single command.

> ## What is a function?
>
> Functions gather a sequence of operations into a whole, preserving it for ongoing use. Functions provide:
>
> * a name we can remember and invoke it by
> * relief from the need to remember the individual operations
> * a defined set of inputs and expected outputs
> * rich connections to the larger programming environment
>
> As the basic building block of most programming languages, user-defined
> functions constitute "programming" as much as any single abstraction can. If
> you have written a function, you are a computer programmer.
{: .callout}

## Defining a function

Let's open a new R script file in the `functions/` directory and call it functions-lesson.R.

```{r}
my_sum <- function(a, b) {
  the_sum <- a + b
  return(the_sum)
}
```

Let’s define a function fahr_to_kelvin that converts temperatures from Fahrenheit to Kelvin:

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

We define `fahr_to_kelvin` by assigning it to the output of `function`. The
list of argument names are contained within parentheses.  Next, the
[body]({{ page.root }}/reference/#function-body) of the function--the statements that are
executed when it runs--is contained within curly braces (`{}`).  The statements
in the body are indented by two spaces.  This makes the code easier to read but
does not affect how the code operates.

When we call the function, the values we pass to it as arguments are assigned to those
variables so that we can use them inside the function.  Inside the function, we
use a [return statement]({{ page.root }}/reference/#return-statement) to send a result back
to whoever asked for it.

Let's try running our function.
Calling our own function is no different from calling any other function:

```{r}
# freezing point of water
fahr_to_kelvin(32)
```

```{r}
# boiling point of water
fahr_to_kelvin(212)
```

> ## Challenge 1
>
> Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin
> and returns that temperature in Celsius
>
> Hint: To convert from Kelvin to Celsius you subtract 273.15
>
> > ## Solution to challenge 1
> >
> > Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin
> > and returns that temperature in Celsius
> >
> > ```{r}
> > kelvin_to_celsius <- function(temp) {
> >  celsius <- temp - 273.15
> >  return(celsius)
> > }
> > ```
> {: .solution}
{: .challenge}

> ## Challenge 2
>
> Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin
> and returns that temperature in Celsius. In addtion, make sure that the function 
> does not produce values smaller than the absolute freeezing point and that we 
> process only numbers (not strings or factors).
>
> Hint: To convert from Kelvin to Celsius you subtract 273.15 (the absolute
> freezing point).
>
> > ## Solution to challenge 2
> >
> > Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin
> > and returns that temperature in Celsius.In addtion, make sure that the function 
> > does not produce values smaller than the absolute freeezing point and that we 
> > process only numbers (not strings).
> >
> > ```{r}
> > kelvin_to_celsius <- function(temp) {
> >   stopifnot(temp < 273,15, is.numeric(temp))
> >   celsius <- temp - 273.15
> >   return(celsius)
> > }
> > ```
> {: .solution}
{: .challenge}

## Combining functions
The real power of functions comes from mixing, matching and combining them
into ever-larger chunks to get the effect we want.

Let's define two functions that will convert temperature from Fahrenheit to
Kelvin, and Kelvin to Celsius:

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}

kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```

> ## Challenge 3
>
> Define the function to convert directly from Fahrenheit to Celsius,
> by reusing the two functions above (or using your own functions if you prefer).
>
>
> > ## Solution to challenge 3
> >
> > Define the function to convert directly from Fahrenheit to Celsius,
> > by reusing these two functions above
> >
> >
> > ```{r}
> > fahr_to_celsius <- function(temp) {
> >   temp_k <- fahr_to_kelvin(temp)
> >   result <- kelvin_to_celsius(temp_k)
> >   return(result)
> > }
> > ```
> {: .solution}
{: .challenge}

> ## Tip
>
> R has some unique aspects that can be exploited when performing
> more complicated operations. We will not be writing anything that requires
> knowledge of these more advanced concepts. In the future when you are
> comfortable writing functions in R, you can learn more by reading the
> [R Language Manual][man] or this [chapter][] from
> [Advanced R Programming][adv-r] by Hadley Wickham. For context, R uses the
> terminology "environments" instead of frames.
{: .callout}

[man]: http://cran.r-project.org/doc/manuals/r-release/R-lang.html#Environment-objects
[chapter]: http://adv-r.had.co.nz/Environments.html
[adv-r]: http://adv-r.had.co.nz/


### Testing and Documenting

Once we start putting things in functions so that we can re-use them, we need
to start testing that those functions are working correctly.  To see how to do
this, let's write a function to center a dataset around a particular value (By
centering a dataset we subtract the men from and add the particular value to
each value of the dataset). For illustration purpose only, we are going to
write both function, the custom function that computes the _mean_, and the
`center` function that makes use of our `custom_mean` function.

```{r}
custom_mean = function(data_vect){
  temp  = sum(data_vect) / length(data_vect)
  return(temp)
}

center <- function(data, desired) {
  new_data <- (data - custom_mean(data)) + desired
  return(new_data)
}
```

We could test this on our actual data, but since we don't know what the values
ought to be, it will be hard to tell if the result was correct.  Instead, let's
create a vector of 0s and then center that around 3. This will make it simple
to see if our function is working as expected:

For the `custom_mean` function:
```{r}
a = c(1, 2, 3)
a
custom_mean(a)
b = c(2, 2, 2)
b
custom_mean(b)
d = c(-1, -2, -3)
d
custom_mean(d)
```

For the `center` function:
```{r}
z <- c(0, 0, 0, 0)
z
center(z, 3)
```

That looks right, so let's try center on our real data. We'll center the
inflammation data from day 4 around 0:

```{r}
dat <- read.csv(file = "data/inflammation-01.csv", header = FALSE)
centered <- center(dat[, 4], 0)
head(centered)
```

It's hard to tell from the default output whether the result is correct, but
there are a few simple tests that will reassure us:

```{r}
# original min
min(dat[, 4])
# centered min
min(centered)
# original mean
mean(dat[, 4])
# centered mean
mean(centered)
# original max
max(dat[, 4])
# centered max
max(centered)
```

That seems almost right: the original mean was about `r round(mean(dat[, 4]),
2)`, so the lower bound from zero is now about `r -round(mean(dat[, 4]), 2)`.
The mean of the centered data is `r mean(centered)`. Other testing options could be.

```{r}
# original standard deviation
sd(dat[, 4])
# centered standard deviation
sd(centered)
```

> 'Whenever you are tempted to type something into a print statement or a debugger
> expression, write it as a test instead.' — Martin Fowler
> [testthat](http://r-pkgs.had.co.nz/tests.html)
{: .callout}

If your project has multiple functions that make use of each other, it is becoming
tedious to rewrite the statements above again and again. Instead, put them
into separate __tests*__ file, the validate your code after you changed
something. The [testthat](http://r-pkgs.had.co.nz/tests.html) provides a
couple of functions to automate and ease the process of testing. Before we start with 
some examples, lets make the advantages of an integrated testing very clear. 

* Code structure- It is easy to write tests for functions that do one
  particular thing. So tests will help you to redefine your problem into single
  bits.
* Confidence about your code- your are more confident about parts of your code
  you wrote tests for.
* Robust code- changes to your code will immediately show failure or success,
  even for parts you did not even think about.
* Increase transferability- running your code on a computer of your colleague
  or on another operating system, your test-suite will show flaws of your
  software. 
* Good entry point to pick up on an older project

But now lets start with some examples and the
[testthat](http://r-pkgs.had.co.nz/tests.html) package.

```{r}
library(testthat)
```

We will go through the two main structures of the package, that are _tests_ and
_expectations_.  _Expectations_,  help you to formalize your interactive tests
and print statements. _Expectations_ are called via the family of functions
that start with `expect_*` and evaluate the properties between __two__ R
objects. Putting our last interactive test into an expectations.

```{r}
expect_equal(sd(centered), sd(dat[, 4]))
```

In case the expression within `expect_that` evaluates to `FALSE`, an error is given.
```{r}
# expect_equal(2, 3)
```

_Tests_, via the `test_that` function on the other site, help us to organize
our expectations by grouping expectations for a single function or an
interaction between multiple functions. That looks like that, if we take the
one of the three interactive tests for our `custom_mean` function.

```{r}
test_that('Testing the custom_mean function', {
  a = c(1, 2, 3)
  expect_equal(custom_mean(a), 2)
})
```
> ## Challenge 4
>
> Write a _test_ as that incoorperates two examples from our `center` function.
>
>
> > ## Solution to challenge 4
> >
> > Write a _test_ as that incoorperates two examples from our `center` function.
> >
> > ```{r}
> > test_that('Testing the center function', {
> >   z <- c(0, 0, 0, 0)
> >   centered = center(z, 3)
> >   expect_equal(centered, rep(3, 4))
> >   # testing the standard deviation
> >   expect_equal(sd(centered), sd(z))
> > })
> > ```
> {: .solution}
{: .challenge}

The __testthat__ package provide as set of functions to run test as convenient
as possible. We will see two examples, namely the `test_file()` and `test_dir()`
functions. The `test_file()` function will run the specified source file. The
functionality is similar to the `source` function, but the output is more
informative. The `test_dir()` on the other hand, will run all files that start
with _test_ in the given directory. Let's put our tests into a file next to our
functions (functions-lesson.R) and see how it works. So we create file called
'test-functions-lesson.R'. The file looks like that:


```{r, eval=F}
# read the file with the functions we want to test
source('functions-lesson.R')

# Test for custom_mean
test_that('Testing the custom_mean function', {
  a = c(1, 2, 3)
  expect_equal(custom_mean(a), 2)
})

# Test for center
test_that('Testing the center function', {
  z <- c(0, 0, 0, 0)
  centered = center(z, 3)
  expect_equal(centered, rep(3, 4))
  # testing the standard deviation
  expect_equal(sd(centered), sd(z))
})
```

Applying the `test_file` function.
```{r, eval=T}
test_file('./functions/test-functions-lesson.R')
```

Applying the `test_dir` function.
```{r, eval=T}
test_dir('./functions')
```
> ## Challenge 5
>
> Extent the file 'test-functions-lesson.R' by writing a test that fails. How
> does the results look like and what does it tell us.
>
> > ## Solution to challenge 5
> >
> > Extent the file 'test-functions-lesson.R' by writing a test that fails. How
> > does the results look like and what does it tell us.
> >
> > ```{r}
> > test_that('Testing the center function', {
> >   z <- c(0, 0, 0, 0)
> >   centered = center(z, 3)
> >   # change the value of the vector to 2, so the lenght is still the same,
> >   # but the value is wrong.
> >   expect_equal(centered, rep(2, 4))
> >   # testing the standard deviation
> >   expect_equal(sd(centered), sd(z))
> > })
> > ```
> {: .solution}
{: .challenge}

> ## Writing Documentation
>
> Formal documentation for R functions is written in separate `.Rd` using a
> markup language similar to [LaTeX][]. You see the result of this documentation
> when you look at the help file for a given function, e.g. `?read.csv`.
> The [roxygen2][] package allows R coders to write documentation alongside
> the function code and then process it into the appropriate `.Rd` files.
> You will want to switch to this more formal method of writing documentation
> when you start writing more complicated R projects.
> 
> It's still possible that our function is wrong, but it seems unlikely enough
> that we should probably get back to doing our analysis.  We have one more task
> first, though: we should write some [documentation]({{ page.root}}/reference#documentation) 
> for our function to remind ourselves later what
> it's for and how to use it.
> 
> A common way to put documentation in software is to add 
> [comments]({{ page.root}}/reference/#comment) like this:
> ```{r}
> center <- function(data, desired) {
>   # return a new vector containing the original data centered around the
>   # desired value.
>   # Example: center(c(1, 2, 3), 0) => c(-1, 0, 1)
>   new_data <- (data - mean(data)) + desired
>   return(new_data)
> }
> ```
> [LaTeX]: http://www.latex-project.org/
> [roxygen2]: http://cran.r-project.org/web/packages/roxygen2/vignettes/rd.html
{: .callout}


### Defining Defaults

We have passed arguments to functions in two ways: directly, as in `mean(dat)`,
and by name, as in `read.csv(file = "data/inflammation-01.csv", header =
FALSE)`. In fact, we can pass the arguments to `read.csv` without naming them:

```{r}
dat <- read.csv("data/inflammation-01.csv", FALSE)
```

However, the position of the arguments matters if they are not named.

```{r, error = TRUE}
dat <- read.csv(FALSE, "data/inflammation-01.csv")
dat <- read.csv(header = FALSE, file = "data/inflammation-01.csv")
```

To understand what's going on, and make our own functions easier to use, let's
re-define our `center` function like this:

```{r}
center <- function(data, desired = 0) {
  # return a new vector containing the original data centered around the
  # desired value (0 by default).
  # Example: center(c(1, 2, 3), 0) => c(-1, 0, 1)
  new_data <- (data - mean(data)) + desired
  return(new_data)
}
```

The key change is that the second argument is now written `desired = 0` instead
of just `desired`.  If we call the function with two arguments, it works as it
did before:

```{r}
test_data <- c(0, 0, 0, 0)
center(test_data, 3)
```

But we can also now call `center()` with just one argument, in which case
`desired` is automatically assigned the default value of `0`:

```{r}
more_data <- 5 + test_data
more_data
center(more_data)
```

This is handy: if we usually want a function to work one way, but occasionally
need it to do something else, we can allow people to pass an argument when they
need to but provide a default to make the normal case easier.

The example below shows how R matches values to arguments

```{r}
display <- function(a = 1, b = 2, c = 3) {
  result <- c(a, b, c)
  names(result) <- c("a", "b", "c")  # This names each element of the vector
  return(result)
}

# no arguments
display()
# one argument
display(55)
# two arguments
display(55, 66)
# three arguments
display (55, 66, 77)
```

As this example shows, arguments are matched from left to right, and any that
haven't been given a value explicitly get their default value. 
We can override this behavior by naming the value as we pass it in:

```{r}
# only setting the value of c
display(c = 77)
```

> ## Matching Arguments
>
> To be precise, R has three ways that arguments are supplied
> by you are matched to the *formal arguments* of the function definition:
>
> 1. by complete name,
> 2. by partial name (matching on initial *n* characters of the argument name), and
> 3. by position.
>
> Arguments are matched in the manner outlined above in *that order*: by
> complete name, then by partial matching of names, and finally by position.
{: .callout}

With that in hand, let's look at the help for `read.csv()`:

```{r, eval=FALSE}
?read.csv
```

There's a lot of information there, but the most important part is the first
couple of lines:

```{r, eval=FALSE}
read.csv(file, header = TRUE, sep = ",", quote = "\"",
         dec = ".", fill = TRUE, comment.char = "", ...)
```

This tells us that `read.csv()` has one argument, `file`, that doesn't have a
default value, and six others that do.
Now we understand why the following gives an error:

```{r, results="hide", error = TRUE}
dat <- read.csv(FALSE, "data/inflammation-01.csv")
```

It fails because `FALSE` is assigned to `file` and the filename is assigned to
the argument `header`.

> ## A Function with Default Argument Values
>
> Rewrite the `rescale` function so that it scales a vector to lie between 0
> and 1 by default, but will allow the caller to specify lower and upper bounds
> if they want.  Compare your implementation to your neighbor's: do the two
> functions always behave the same way?
>
> > ## Solution
> > ~~~
> > rescale <- function(v, lower = 0, upper = 1) {
> >   # Rescales a vector, v, to lie in the range lower to upper.
> >   L <- min(v)
> >   H <- max(v)
> >   result <- (v - L) / (H - L) * (upper - lower) + lower
> >   return(result)
> > }
> > ~~~
> > {: .r}
> {: .solution}
{: .challenge}

```{r, include=FALSE}
rescale <- function(v, lower = 0, upper = 1) {
  # Rescales a vector, v, to lie in the range lower to upper.
  L <- min(v)
  H <- max(v)
  result <- (v - L) / (H - L) * (upper - lower) + lower
  return(result)
}
answer <- rescale(dat[, 4], lower = 2, upper = 5)
min(answer)
max(answer)
answer <- rescale(dat[, 4], lower = -5, upper = -2)
min(answer)
max(answer)
```
